`mathml`: Translate R Expressions to MathML and LaTeX

### 2023-06-18 rj/mathml.Rmd
Reviewer 3

The R\ extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and LaTeX.
Since recently [@R, version 4.2], R\'s manual pages include support for
mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a big
improvement. However, except for special cases such as regression
models [@equatiomatic] and R's own plotmath annotation, rules for the mapping of
built-in language elements to their mathematical representation are still
lacking. So far, R\ expressions such as `pbinom(k, N, p)` are printed as they
are; pretty mathematical formulae such as \(P_{\mathrm{Bi}}(X \le k; N, p)\)
require explicit LaTeX commands
like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
cases, these commands are tedious to type in and their source code is hard to
read.

The present R\ package defines a set of rules for the automatic translation of
R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and potentially reduces mistakes, as will be illustrated below.

@@

````{r}
library(mathml)
mathjax(term)
````

Some of the curly braces are not really needed in this simple example, but are
necessary in edge cases. We can include the output in a R\ Markdown document by
specifying `results='asis'` in the R\ code chunk of R\ Markdown, as is done in all
examples below. The R\ function `mathout()` has been defined in the preamble of
this article; it wraps a call to `mathml()` for HTML output and `mathjax()` for
LaTeX output.

````{r, results='asis'}
mathout(term)
````

For convenience, the translation of the R expressions is achieved through a
Prolog interpreter provided by another package 
R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic programming
language with many applications in expert systems, computer linguistics and
symbolic artificial intelligence. The strength of Prolog lies in its concise
representation of facts and rules for knowledge and grammar, as well as its
efficient built-in search engine for closed world domains. Whereas Prolog is
weak in statistical computation, but strong in symbolic manipulation, the
converse may be said for the R\ language. \pkg{rolog} bridges this gap by
providing an interface to a SWI-Prolog distribution [@swipl] in R. The
communication between the two systems is mainly in the form of queries from
R\ to Prolog, but two Prolog functions allow to ring back and evaluate terms
in R.

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the section on errors below).

@@

>>>>

The R\ extension of the markdown language [@Xie2018;rmarkdown] enables
reproducible statistical reports with nice typesetting in HTML, Microsoft Word,
and LaTeX. Since recently [@R, version 4.2], R\'s manual pages include support
for mathematical expressions [@Sarkar2022;@Viechtbauer2022], which already is a
big improvement. However, except for special cases such as regression
models [@equatiomatic] and R's own plotmath annotation, rules for the mapping of
built-in language elements to their mathematical representation are still
lacking. So far, R\ expressions such as `pbinom(k, N, p)` are printed as they
are; pretty mathematical formulae such as \(P_{\mathrm{Bi}}(X \le k; N, p)\)
require explicit LaTeX commands
like `P_{\mathrm{Bi}}\left(X \le k; N, p\right)`. Except for very basic use
cases, these commands are tedious to type in and their source code is hard to
read.

The present R\ package defines a set of rules for the automatic translation of
R\ expressions to mathematical output in R\ Markdown documents [@Xie2020] and
Shiny apps [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R\ elements by custom mathematical
signs.

The main feature of the package is that the same R\ expressions and equations
can be used for both mathematical typesetting and calculations. This saves time
and potentially reduces mistakes, as will be illustrated below. Readers should
have basic knowledge of knitr and R\ Markdown to be able to follow this
article [@knitr;@rmarkdown]. To extend and customize the package, some basic
knowledge of Prolog is needed.

@@

````{r}
library(mathml)
substr(mathml(term), 1, 70)
mathjax(term)
````

Some of the curly braces are not really needed in the LaTeX output, but are
necessary in edge cases. The package also includes a function `mathout()` that
wraps a call to `mathml()` for HTML output and `mathjax()` for LaTeX output.
Moreover, the function `math(x)` adds the class `"math"` to its argument, such
that a special knitr printing function is invoked. We can include the output in
a R\ Markdown document by specifying `results="asis"` in the R\ code chunk of
R\ Markdown, as is done in all examples below.

``````markdown
`r ''`````{r, results="asis"}
math(term)
````
``````

````{r, results="asis", echo=FALSE}
math(term)
````

Similarly, `inline()` produces inline
output, `` `r "\x60r inline(term)\x60"` `` yields `r inline(term)`.

@@

## Implementation details

For convenience, the translation of the R expressions is achieved through a
Prolog interpreter provided by another R\ package \CRANpkg{rolog} [@rolog]. If a
version of SWI-Prolog [@swipl] is found on the system, \CRANpkg{rolog} connects
to it. Alternatively, the SWI-Prolog runtime libraries can be conveniently
accessed by installing the R\ package \CRANpkg{rswipl} [@rswipl]. Prolog is a
classical logic programming language with many applications in expert systems,
computer linguistics and symbolic artificial intelligence. The strength of
Prolog lies in its concise representation of facts and rules for knowledge and
grammar, as well as its efficient built-in search engine for closed world
domains. Whereas Prolog is weak in statistical computation, but strong in
symbolic manipulation, the converse may be said for the R\ language. \pkg{rolog}
bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two Prolog functions allow to ring
back and evaluate terms in R.

The relevant rule for `ml/3` builds the MathML entity `mrow([N, mo(&(af)), X])`,
with `N` representing the name of the function and `X` its arguments and
parameters enclosed in parentheses. A corresponding rule `jax/3` does the same
for MathJax/LaTeX. A list of flags can be used for context-sensitive
translation (see, e.g., the section on errors above).
